\bf Mechanistic models for panel data: analysis of ecological experiments with four interacting species

### 2024-03-20 ms.Rnw
Update

<<Setup, include=FALSE,echo=FALSE,results=hide>>=
 library(plyr)
 library(reshape2)
library(patchwork)
 library(magrittr)
 library(foreach)
 library(readxl)
 library(doParallel)
 registerDoParallel(cores=36)
 library(pomp)
 library(panelPomp)
 library(ggplot2)

@@

\bigskip
\begin{abstract}
In ecological research, panel data, often comprising multi-magnitude time series, presents unique analytical challenges due to its high dimensionality. This study employs the PanelPOMP framework and the Panel Iterated Filtering (PIF) method to explore the practicality and advancement of mechanistic model development for such complex datasets. Mechanistic models, especially valuable in ecological research, allow for quantitative analysis and deeper understanding of dynamic systems. However, they are often constrained by the non-Gaussian nature of ecological data. Our study applies these methods to the Daphnia panel data, consisting of independent time series with 10 data points each, to derive parameters that minimize the Akaike Information Criterion (AIC). The comparison of AIC values across models reveals the significant impact of alga, the prey of Daphnia, and parasite on their population dynamics which is observed in the SIRPF model highlights this relationship. The panelPOMP framework's ability to handle complex panel data analysis is demonstrated, emphasizing its utility in uncovering latent processes not captured by traditional models. However, the study also discusses limitations and interpretations of mechanistic models and PanelPOMP framework in ecological data analysis.
\end{abstract}

@@

In this research, we will show the advancement of the panelPOMP model by analyzing the Daphnia panel data generated in the experiment of Searle et al. \citep{Searle2016}. The research group of Searle et al. collected panel data on the population densities of different species of Daphnia under different conditions, and interpret how invasive Daphnia species (\textit{lumholtzi}) affect the response of native Daphnia species (\textit{dentifera}) to parasites with panel data \citep{Searle2016}. In their lab experiment, they generated standard ecological panel data featuring various independent trials under the same experimental conditions, with each consisting of a brief, nonlinear, and nonstationary time series. Conventional mechanistic theories cannot directly interpret these observations. Utilizing the panelPOMP approach allows for the combination of theoretical models and empirical data, enabling the extraction of the comprehensive information found in each individual time series \citep{Carles2020}.\\

For a more in-depth examination of the concealed processes in the panel data, this research incorporates the Susceptible-Infected-Recovered-Food-Parasite (SIRPF) model - within the panelPOMP framework. After applying the model to the Daphnia panel data, the maximum log-likelihood and best parameter estimates are determined via an iterative filtering approach. These findings are subsequently compared to discover a more accurate and appropriate model for the given data. The superior performance of the SIRPF model implies that the population density of alga \textit{Ankistrodesmus falcatus}, which is Daphnia's prey, impact the dynamics of both species of Daphnia. It is worth noting that the population density of alga is neither observable nor recorded in the experiment. That implies the mechanistic models under panelPOMP potentially better explain the latent process and show the importance of unobservable processes \citep{Edwards2012}. At the same time, the panelPOMP framework provides an estimation of the parameters in the model, which helps to understand the hidden mechanisms in the ecological system \citep{Carles2020}. Based on the estimation, a detailed interpretation of relationship between invasive host and native host with presence of parasite is explaned.\\

Our objective is to evaluate the explanatory capabilities of two traditional ecological models when applied to the Daphnia panel data, and to estimate the parameter sets for a more accurate depiction of the hidden mechanisms within the ecological system. By analyzing Daphnia panel data, we hope to show the limitations and interpretations of mechanistic models in dealing with ecological data. Additionally, the later section will further discuss how the panelPOMP framework and PIF contribute to overcoming the inherent challenges associated with fitting ecological models.

@@

The dynamics of our study are derived from the Searle's comparison experiment, which focuses on two Daphnia species: Daphnia \textit{dentifera Forbes}, natively prevalent in North American stratified lakes \citep{searle2016salinization}, and Daphnia \textit{lumholtzi Sars}, an invasive species originating from Africa, Asia, and Australia. D. lumholtzi, first observed in North America in the 1990s, has rapidly expanded across the United States, impacting native species and ecosystem dynamics \citep{havel1993daphnia}. These Daphnia species, referred to as the native and invasive hosts, are bred in controlled laboratory environments using isofemale lines. For their experiments, six clones of each species were selected to eliminate selection bias \citep{Searle2016}.\\

The dynamics also involves \textit{Metschnikowia bicuspidata}, a parasitic fungus predominantly infecting the native host. This parasite, ingested as spores by Daphnia, aggressively affects the host's fecundity and longevity \citep{duffy2010temporal}. The parasite's presence is identifiable by the opaqueness it causes in normally transparent hosts, with spore release upon host death. Searle's study utilized a M. bicuspidata strain from Baker Lake, Michigan, with consistent infectivity and virulence characteristics in different samples \citep{Searle2016}. The final species in our study is \textit{Ankistrodesmus falcatus}, a green alga that serves as the primary food source for Daphnia. \textit{Ankistrodesmus falcatus} is a globally distributed freshwater chlorophyte, is known for its role in aquatic communities and has been utilized in ecological research due to its nutritional properties \citep{arnold1971ingestion}. Its significance is particularly evident in controlled experimental settings where the growth conditions and feeding rates of Daphnia species can be closely monitored and manipulated \citep{martinez1994effect}.\\

Drawing upon the empirical findings from Searle's laboratory, it was observed that while the two Daphnia host species exhibited comparable resource acquisition rates, the invasive species demonstrated a more rapid reproductive cycle \citep{Searle2016}. This observation suggests a potential competitive advantage for the invasive species. Building on these insights, Searle and colleagues structured the ecological system we analyze here, focusing on dynamics at the community level. Their approach involved a mesocosm experiment that varied species composition and parasite presence. The experimental design was a 3x2x1 factorial arrangement, incorporating three species combinations (native only, invasive only, or both existing species), two parasite exposure scenarios (exposed or unexposed), and a single algal food source. This design enabled an in-depth analysis of the invasive host's impact on native populations, manifesting either directly through competition for resources, indirectly via the amplification of a mutual parasite, or through a reduction in parasite effects.\\

In essence, the impact of an invasive Daphnia \textit{lumholtzi}, particularly in terms of resource competition like food, can be significant. This scenario would lead to a decreased survival rate of the native host in the presence of the invasive species, irrespective of parasite interference. This effect is quantifiable through the ratio of natality to mortality in the native host population. Conversely, should the invasive species primarily influence the native host indirectly via interactions with parasites, a different pattern emerges. Here, the survivability of the native host is expected to be lower in scenarios involving both the parasite and the invasive host, as opposed to situations where the native host is only affected by the parasite. This effect is quantifiable through the ratio of natality to mortality in the native host population as well.\\

\subsection{Non-parasitized Dynamics}
This research stratifies the ecological dynamics above into two analytical categories delineated by parasitic presence. Under the 'Non-parasitized Dynamics' classification, three experimental groups of Daphnia species—solely native, solely invasive, or a combined population—are statistically analyzed. This categorization facilitates a quantitative assessment of the interaction dynamics between native and invasive Daphnia species, particularly in environments not influenced by parasitic interactions.\\

In the Non-parasitized level, indoor mesocosms were used \citep{Searle2016}, each comprising a 18.9 L bucket with 15 L of high-hardness COMBO media. The experimental design involved 10 replicates for single-species treatments, and 9 replicates for mixed-species treatment. A nitrogen-to-phosphorus ratio of 25:1 was established, and each mesocosm was seeded with 2.5 $\times$ $10^8$ cells of \textit{Ankistrodesmus falcatus}. The laboratory conditions included an average temperature of 23.27$^\circ C$ ($\pm$0.2) and a 16-hour light, 8-hour dark cycle. For single-species treatments, either 45 native or invasive Daphnia hosts were added, representing six genotypes per species to eliminate the selection bias. In mixed-species treatments, 35 native and 10 invasive hosts were introduced, simulating early invasion stages.\\

\subsection{Parasitized Dynamics}
Parasitized dynamics is employed with a key difference, which is the presence of the parasite \textit{Metschnikowia bicuspidata} . Based on the same setup of Non-parasitized Dynamics, after a 4-day acclimation period in the mesocosms, hosts are exposed to parasites, with spore counts set at 25 per mL in parasite-exposed treatments. This category allows for the investigation of the impact of parasites on the interaction dynamics between native and invasive Daphnia species, particularly how parasitism influences competition and survival rates. This approach complements the "Non-parasitized Dynamics" by providing a contrasting scenario where parasitic factors are at play, thereby enriching the understanding of ecological interactions under the presence of parasite.\\

\subsection{Data Collection and Visualization}
Population densities and infection prevalence(if exists) were quantified every five days, beginning a week after the experiment started. On each sampling day, mesocosms were thoroughly mixed, and a 1-liter sample was removed and replaced with fresh COMBO media. The samples were concentrated using a 153-$\mu m$ mesh and the Daphnia were counted under a microscope. For each Daphnia, species (native or invasive), infection status, age, and sex were determined. Regular stirring of mesocosms ensured resuspension of algae and parasites. Algal and nutrient levels were supplemented bi-weekly, with COMBO media added as necessary. The experiment concluded after 52 days, totaling 10 sampling sessions \citep{Searle2016}.\\

As juveniles rarely exhibit clear signs of infection, only adult Daphnia were used for quantifying effects in both Non-parasitized and Parasitized Dynamics to minimize information bias. Figures below illustrates these dynamics, with the figure 1 depicting Non-parasitized Dynamics (mixed-species) where changes in native and invasive host densities reflect direct resource competition, for example, food alga \textit{Ankistrodesmus falcatus}. Figure 2 shows Parasitized Dynamics, where the interplay of susceptible and infected Daphnia of different species is further influenced by parasites, complicating the variation of susceptible and infected Daphnia of different races, which makes it difficult to directly analyze the sources of influence on their variation. Non-parasitized Dynamics will be analyzed at the first place to help understand the effect of resource competition in mixed-species. And based on this result, the correlation between native host and invasive host in Parasitized Dynamics will be further explored.

@@

   scale_x_continuous(limits = c(0, 52)) +
   scale_y_continuous(limits = c(0, 175)) +
   facet_grid(. ~ group, scales = "fixed", space = "free") +
   theme(
     strip.text = element_text(size=10),
     axis.text.x=element_text(color="black"),
     axis.ticks.x=element_line(color="black"),
     axis.text.y=element_text(color="black"),
     axis.ticks.y=element_line(color="black"),
     axis.title.x=element_blank(),
     axis.title.y=element_blank(),
     legend.position="none",
     panel.spacing = unit(0, "lines"), # Adjust spacing between panels
     strip.background = element_blank(), # Optional: Remove facet label background
     strip.placement = "outside"
   ) +
   labs(title="Daphnia density in Parasitized Dynamics") +
   guides(colour=guide_legend(title="Species"))
combined_plot <- p1 / p2 
combined_plot
@
\caption{Plot of density of hosts in dynamics including both native and invasive host species of Daphnia. The trail A to J represents different buckets in different experiment. The black lines represent the Daphnia dentifera density, the red lines show the Daphnia lumholtzi density, the green lines are the infected Daphnia lumholtzi density, and the blue lines indicate the infected Daphnia dentifera density.}
\end{figure}

@@

To analyze the dynamics of Parasitized and Non-parasitized systems, the Susceptible-Infected-Recovered-Food-Parasite (SIRPF) model, an adaptation of the SIR model \citep{weiss2013sir}, is utilized. This model integrates the aspects of food and parasites, unlike the classical SIR model, to more accurately reflect the interactions between native and invasive hosts in the context of resource availability and parasitic influence. The SIRPF model incorporates these factors as quantitative latent state variables, enabling a more comprehensive understanding of the system's dynamics. In the meantime, the ecological processes of Daphnia can be viewed as a Markov process, where the population of Susceptible and Infected individuals is observable, while the amount of food and parasites represent unobservable latent states. Utilizing panel data on Daphnia, the PanelPOMP framework becomes a valuable analytical tool \citep{Carles2020}. Despite the lack of direct measurement data for these latent states, PanelPOMP, which has proven efficiency in analyzing ecological panel data, facilitates a robust analysis of these dynamics, effectively handling the complexities inherent in ecological systems.\\

@@

\draw [->, bendy] (S) to  (I);
\draw [->, bendy] (I) to  (S);
\draw [->, bendy] (P) to (I);
% \path [->] (S) edge (I) 
%            (P) edge (I);
\draw [->, bendy] (I) to (P);
\end{tikzpicture}
}
\end{center}
\vspace{-5mm}
\caption{A flow diagram for the SIRPF model. Apart from the quantified states above, the graph includes $R$ state, which represents deadth status. For Daphnia either dentifera or lumholtzi, susceptible, infected Daphnia and food retain the ability to produce offspring, which is represented in the graph as a self-recycling of $S$ and $F$, as well as transaction from $I$ to $S$. Some susceptible Daphnia get infected by parasites, transitioning from $S$ to $I$. Both susceptible and infected Daphnia contribute to the $R$ state upon death, releasing surviving parasites back into the environment. And the parasites continue to impact other individuals and get into another cycle. Food is consumed by both $S$ and $I$ individuals. Over time, elements in $F$ and $P$ also transit to the $R$ state.}
\label{fig:flow_SIRPF}
\end{figure}

@@

Each equation intricately ties the state variables to depict the complex dynamics of the single species Parasitized system. Building on the SIRPF model, the SIRPF2 model further adapts to a mixed-race Parasitized system by introducing another Daphnia species. This advanced model incorporates both invasive($S_i , I_i$) and native($S_n , I_n$)hosts, along with food($F$) and parasites($P$), and delineates their specific interactions within the ecosystem as the following graph and model shows. In the SIRPF2 model, the subscript n is used to denote parameters and variables related to the native host, i for those pertaining to the invasive host, and p for the parasite spores. The SIRPF2 model thus offers a more detailed representation of the ecological dynamics by accounting for the complex interplay between different species and environmental factors in a mechanistic framework.
\begin{figure}
\begin{center}
%%%%% SEAIR diagram
\resizebox{8cm}{!}{
\begin{tikzpicture}[
  square/.style={rectangle, draw=black, minimum width=0.5cm, minimum height=0.5cm, rounded corners=.1cm, fill=blue!8},
  rhombus/.style={diamond, draw=black, minimum width=0.1cm, minimum height=0.1cm, fill=purple!8,aspect = 1},
  travel/.style={circle, draw=black, minimum width=0.5cm, minimum height=0.5cm, fill=green!8},
  report/.style={shape=regular polygon, regular polygon sides=8, draw, fill=red!8,minimum size=0.6cm,inner sep=0cm},
  bendy/.style={bend left=10},
  bendy2/.style={bend left=100},
  bendy3/.style={bend left=-100},
  >/.style={shorten >=0.25mm}, % redefine arrow to stop short of node
  >/.tip={Stealth[length=1.5mm,width=1.5mm]} % redefine arrow style
]
\tikzset{>={}}; % this is needed to implement the arrow redefinition

@@

\end{tikzpicture}
}
\end{center}
\vspace{-5mm}
\caption{A flow diagram for the SIRPF2 model. Apart from the quantified states above, the graph includes $R$ state, which represents deadth status. The flow diagram shows the how two species of Daphnia interact with each other indirectly through $F$ and $P$. The marginal processes have been shown in SIRPF mode, and SIRPF2 model is the adaption form in mixed-species dynamcis}
\label{fig:flow_SIRPF2}
\end{figure}

@@

In the SIRPF2 model, detailed in Table 1, parameters and variables are distinctly annotated for native hosts (subscript $n$), invasive hosts (subscript $i$), and parasite spores (subscript $p$). Intraspecific competition among hosts is conceptualized via the Lotka-Volterra framework, with $r_n$ and $r_i$ denoting respective reproductive rates. Host density growth, influenced by algal density, is posited to be linear, indicating a direct relationship between food availability and Daphnia reproduction. Mortality rates of susceptible hosts are denoted as $\theta_{S_n}$ and $\theta_{S_i}$. The model encapsulates the dynamics of infection through \textit{Metschnikowia bicuspidata} intake, affecting susceptible hosts. This interaction consequently modulates reproductive rates, represented as $\xi r_n$. Infection rates correlate with spore density ($P$) and host ingestion rates ($f_{S_n}$, $f_{S_i}$). Upon death, infected hosts ($\theta_{I_n}$, $\theta_{I_i}$) release spores, which are subsequently subject to ingestion and environmental degradation. The consumption dynamics of food by susceptible and infected hosts are captured by $\gamma_n$ and $\gamma_i$, and food growth is inversely proportional to its carrying capacity ($K_F$), reflective of resource limitation. The model also incorporates a constant removal rate $\delta$, representing the loss of hosts, spores, and food due to external sampling. The constraint of food supply, indicated by a constant rate $\mu = 0.37 \times 10^5$ cells per day, is based on observed declines in Daphnia populations in the panel data, suggesting a pivotal role of food availability in driving population dynamics.\\

@@

\draw [dotted, thick, red] (3.5, -0.75) rectangle (6.5, 1.5);
\end{tikzpicture}
}
\end{center}
\vspace{-5mm}
\caption{A flow diagram for the SF and SF2 model. In this system, both types of Daphnia and the food resource contribute to reproduction, depicted as self-recycling loops for $S_n$, $S_i$, and $F$. As time progresses, individuals in $F$, $S_n$, and $S_i$ transition to the $R$ state, signifying mortality. While the whole graph show SF2 model, if we remove the node in the dashed rectangle, the graph will represent the SF model.}
\label{fig:SF&SF2}
\end{figure}

@@

The SIRPF model is a powerful tool for analyzing ecological data, as it can be used to estimate parameters and predict future population dynamics. The model's flexibility allows for the incorporation of a variety of factors, including food resources and parasitism, that can be used to assess the impact of these factors on the population dynamics of Daphnia. The model's ability to estimate parameters and predict future population dynamics is a powerful tool for analyzing ecological data. The model's flexibility allows for the incorporation of a variety of factors, including food resources and parasitism, that can be used to assess the impact of these factors on the population dynamics of Daphnia.\\

>>>>

<<Setup, include=FALSE,echo=FALSE,results=hide>>=
 library(plyr)
 library(reshape2)
library(patchwork)
 library(magrittr)
 library(foreach)
 library(readxl)
 library(doParallel)
 registerDoParallel(cores=36)
 library(pomp)
 library(panelPomp)
 library(ggplot2)
 library(gridExtra)

@@

\bigskip
\begin{abstract}
In ecological research, panel data, often comprising multi-magnitude time series, presents unique analytical challenges due to its high dimensionality. This study employs the PanelPOMP framework and the Panel Iterated Filtering (PIF) method to explore the practicality and advancement of mechanistic model development for such complex datasets. Mechanistic models, especially valuable in ecological research, allow for quantitative analysis and deeper understanding of dynamic systems. However, they are often constrained by the non-Gaussian nature of ecological data. Our study applies these methods to the \textit{Daphnia} panel data, consisting of independent time series with 10 data points each, to derive parameters that minimize the Akaike Information Criterion (AIC). The comparison of AIC values across models reveals the significant impact of alga, the prey of \textit{Daphnia}, and parasite on their population dynamics which is observed in the SIRPF model highlights this relationship. The panelPOMP framework's ability to handle complex panel data analysis is demonstrated, emphasizing its utility in uncovering latent processes not captured by traditional models. However, the study also discusses limitations and interpretations of mechanistic models and PanelPOMP framework in ecological data analysis.
\end{abstract}

@@

In this research, we will show the advancement of the panelPOMP model by analyzing the \textit{Daphnia} panel data generated in the experiment of Searle et al. \citep{Searle2016}. The research group of Searle et al. collected panel data on the population densities of different species of \textit{Daphnia} under different conditions, and interpret how invasive \textit{Daphnia} species (\textit{lumholtzi}) affect the response of native \textit{Daphnia} species (\textit{dentifera}) to parasites with panel data \citep{Searle2016}. In their lab experiment, they generated standard ecological panel data featuring various independent trials under the same experimental conditions, with each consisting of a brief, nonlinear, and nonstationary time series. Conventional mechanistic theories cannot directly interpret these observations. Utilizing the panelPOMP approach allows for the combination of theoretical models and empirical data, enabling the extraction of the comprehensive information found in each individual time series \citep{Carles2020}.\\

For a more in-depth examination of the concealed processes in the panel data, this research incorporates the Susceptible-Infected-Recovered-Food-Parasite (SIRPF) model - within the panelPOMP framework. After applying the model to the \textit{Daphnia} panel data, the maximum log-likelihood and best parameter estimates are determined via an iterative filtering approach. These findings are subsequently compared to discover a more accurate and appropriate model for the given data. The superior performance of the SIRPF model implies that the population density of alga \textit{Ankistrodesmus falcatus}, which is \textit{Daphnia}'s prey, impact the dynamics of both species of \textit{Daphnia}. It is worth noting that the population density of alga is neither observable nor recorded in the experiment. That implies the mechanistic models under panelPOMP potentially better explain the latent process and show the importance of unobservable processes \citep{Edwards2012}. At the same time, the panelPOMP framework provides an estimation of the parameters in the model, which helps to understand the hidden mechanisms in the ecological system \citep{Carles2020}. Based on the estimation, a detailed interpretation of relationship between invasive host and native host with presence of parasite is explaned.\\

Our objective is to evaluate the explanatory capabilities of two traditional ecological models when applied to the \textit{Daphnia} panel data, and to estimate the parameter sets for a more accurate depiction of the hidden mechanisms within the ecological system. By analyzing \textit{Daphnia} panel data, we hope to show the limitations and interpretations of mechanistic models in dealing with ecological data. Additionally, the later section will further discuss how the panelPOMP framework and PIF contribute to overcoming the inherent challenges associated with fitting ecological models.

@@

The dynamics of our study are derived from the Searle's comparison experiment, which focuses on two \textit{Daphnia} species: \textit{Daphnia} \textit{dentifera Forbes}, natively prevalent in North American stratified lakes \citep{searle2016salinization}, and \textit{Daphnia} \textit{lumholtzi Sars}, an invasive species originating from Africa, Asia, and Australia. D. lumholtzi, first observed in North America in the 1990s, has rapidly expanded across the United States, impacting native species and ecosystem dynamics \citep{havel1993daphnia}. These \textit{Daphnia} species, referred to as the native and invasive hosts, are bred in controlled laboratory environments using isofemale lines. For their experiments, six clones of each species were selected to eliminate selection bias \citep{Searle2016}.\\

The dynamics also involves \textit{Metschnikowia bicuspidata}, a parasitic fungus predominantly infecting the native host. This parasite, ingested as spores by \textit{Daphnia}, aggressively affects the host's fecundity and longevity \citep{duffy2010temporal}. The parasite's presence is identifiable by the opaqueness it causes in normally transparent hosts, with spore release upon host death. Searle's study utilized a M. bicuspidata strain from Baker Lake, Michigan, with consistent infectivity and virulence characteristics in different samples \citep{Searle2016}. The final species in our study is \textit{Ankistrodesmus falcatus}, a green alga that serves as the primary food source for \textit{Daphnia}. \textit{Ankistrodesmus falcatus} is a globally distributed freshwater chlorophyte, is known for its role in aquatic communities and has been utilized in ecological research due to its nutritional properties \citep{arnold1971ingestion}. Its significance is particularly evident in controlled experimental settings where the growth conditions and feeding rates of \textit{Daphnia} species can be closely monitored and manipulated \citep{martinez1994effect}.\\

Drawing upon the empirical findings from Searle's laboratory, it was observed that while the two \textit{Daphnia} host species exhibited comparable resource acquisition rates, the invasive species demonstrated a more rapid reproductive cycle \citep{Searle2016}. This observation suggests a potential competitive advantage for the invasive species. Building on these insights, Searle and colleagues structured the ecological system we analyze here, focusing on dynamics at the community level. Their approach involved a mesocosm experiment that varied species composition and parasite presence. The experimental design was a 3x2x1 factorial arrangement, incorporating three species combinations (native only, invasive only, or both existing species), two parasite exposure scenarios (exposed or unexposed), and a single algal food source. This design enabled an in-depth analysis of the invasive host's impact on native populations, manifesting either directly through competition for resources, indirectly via the amplification of a mutual parasite, or through a reduction in parasite effects.\\

In essence, the impact of an invasive \textit{D. lumholtzi}, particularly in terms of resource competition like food, can be significant. This scenario would lead to a decreased survival rate of the native host in the presence of the invasive species, irrespective of parasite interference. This effect is quantifiable through the ratio of natality to mortality in the native host population. Conversely, should the invasive species primarily influence the native host indirectly via interactions with parasites, a different pattern emerges. Here, the survivability of the native host is expected to be lower in scenarios involving both the parasite and the invasive host, as opposed to situations where the native host is only affected by the parasite. This effect is quantifiable through the ratio of natality to mortality in the native host population as well.\\

\subsection{Non-parasitized Dynamics}
This research stratifies the ecological dynamics above into two analytical categories delineated by parasitic presence. Under the 'Non-parasitized Dynamics' classification, three experimental groups of \textit{Daphnia} species—solely native, solely invasive, or a combined population—are statistically analyzed. This categorization facilitates a quantitative assessment of the interaction dynamics between native and invasive \textit{Daphnia} species, particularly in environments not influenced by parasitic interactions.\\

In the Non-parasitized level, indoor mesocosms were used \citep{Searle2016}, each comprising a 18.9 L bucket with 15 L of high-hardness COMBO media. The experimental design involved 10 replicates for single-species treatments, and 9 replicates for mixed-species treatment. A nitrogen-to-phosphorus ratio of 25:1 was established, and each mesocosm was seeded with 2.5 $\times$ $10^8$ cells of \textit{Ankistrodesmus falcatus}. The laboratory conditions included an average temperature of 23.27$^\circ C$ ($\pm$0.2) and a 16-hour light, 8-hour dark cycle. For single-species treatments, either 45 native or invasive \textit{Daphnia} hosts were added, representing six genotypes per species to eliminate the selection bias. In mixed-species treatments, 35 native and 10 invasive hosts were introduced, simulating early invasion stages.\\

\subsection{Parasitized Dynamics}
Parasitized dynamics is employed with a key difference, which is the presence of the parasite \textit{Metschnikowia bicuspidata} . Based on the same setup of Non-parasitized Dynamics, after a 4-day acclimation period in the mesocosms, hosts are exposed to parasites, with spore counts set at 25 per mL in parasite-exposed treatments. This category allows for the investigation of the impact of parasites on the interaction dynamics between native and invasive \textit{Daphnia} species, particularly how parasitism influences competition and survival rates. This approach complements the "Non-parasitized Dynamics" by providing a contrasting scenario where parasitic factors are at play, thereby enriching the understanding of ecological interactions under the presence of parasite.\\

\subsection{Data Collection and Visualization}
Population densities and infection prevalence(if exists) were quantified every five days, beginning a week after the experiment started. On each sampling day, mesocosms were thoroughly mixed, and a 1-liter sample was removed and replaced with fresh COMBO media. The samples were concentrated using a 153-$\mu m$ mesh and the \textit{Daphnia} were counted under a microscope. For each \textit{Daphnia}, species (native or invasive), infection status, age, and sex were determined. Regular stirring of mesocosms ensured resuspension of algae and parasites. Algal and nutrient levels were supplemented bi-weekly, with COMBO media added as necessary. The experiment concluded after 52 days, totaling 10 sampling sessions \citep{Searle2016}.\\

As juveniles rarely exhibit clear signs of infection, only adult \textit{Daphnia} were used for quantifying effects in both Non-parasitized and Parasitized Dynamics to minimize information bias. Figures below illustrates these dynamics, with the figure 1 depicting Non-parasitized Dynamics (mixed-species) where changes in native and invasive host densities reflect direct resource competition, for example, food alga \textit{Ankistrodesmus falcatus}. Figure 2 shows Parasitized Dynamics, where the interplay of susceptible and infected \textit{Daphnia} of different species is further influenced by parasites, complicating the variation of susceptible and infected \textit{Daphnia} of different races, which makes it difficult to directly analyze the sources of influence on their variation. Non-parasitized Dynamics will be analyzed at the first place to help understand the effect of resource competition in mixed-species. And based on this result, the correlation between native host and invasive host in Parasitized Dynamics will be further explored.

@@

   scale_x_continuous(limits = c(0, 52)) +
   scale_y_continuous(limits = c(0, 175)) +
   facet_grid(. ~ group, scales = "fixed", space = "free") +
   theme(
     strip.text = element_text(size=10),
     axis.text.x=element_text(color="black"),
     axis.ticks.x=element_line(color="black"),
     axis.text.y=element_text(color="black"),
     axis.ticks.y=element_line(color="black"),
     axis.title.x=element_blank(),
     axis.title.y=element_blank(),
     legend.position="none",
     panel.spacing = unit(0, "lines"), # Adjust spacing between panels
     strip.background = element_blank(), # Optional: Remove facet label background
     strip.placement = "outside"
   ) +
   labs(title="Daphnia density in Parasitized Dynamics") +
   guides(colour=guide_legend(title="Species"))
grid.arrange(p1, p2, ncol = 1)
@
\caption{Plot of density of hosts in dynamics including both native and invasive host species of \textit{Daphnia}. The trail A to J represents different buckets in different experiment. The black lines represent the \textit{D. dentifera} density, the red lines show the \textit{D. lumholtzi} density, the green lines are the infected \textit{D. dentifera} density, and the blue lines indicate the infected \textit{D. lumholtzi} density.}
\end{figure}

@@

To analyze the dynamics of Parasitized and Non-parasitized systems, the Susceptible-Infected-Recovered-Food-Parasite (SIRPF) model, an adaptation of the SIR model \citep{weiss2013sir}, is utilized. This model integrates the aspects of food and parasites, unlike the classical SIR model, to more accurately reflect the interactions between native and invasive hosts in the context of resource availability and parasitic influence. The SIRPF model incorporates these factors as quantitative latent state variables, enabling a more comprehensive understanding of the system's dynamics. In the meantime, the ecological processes of \textit{Daphnia} can be viewed as a Markov process, where the population of Susceptible and Infected individuals is observable, while the amount of food and parasites represent unobservable latent states. Utilizing panel data on \textit{Daphnia}, the PanelPOMP framework becomes a valuable analytical tool \citep{Carles2020}. Despite the lack of direct measurement data for these latent states, PanelPOMP, which has proven efficiency in analyzing ecological panel data, facilitates a robust analysis of these dynamics, effectively handling the complexities inherent in ecological systems.\\

@@

\draw [->, bendy] (S) to  (I);
\draw [->, bendy] (I) to  (S);
\draw [->, bendy] (P) to (I);
% \path [->] (S) edge (I) 
%            (P) edge (I);
\draw [->, bendy] (I) to (P);
\end{tikzpicture}
}
\end{center}
\vspace{-5mm}
\caption{A flow diagram for the SIRPF model. Apart from the quantified states above, the graph includes $R$ state, which represents deadth status. For \textit{Daphnia} either dentifera or lumholtzi, susceptible, infected \textit{Daphnia} and food retain the ability to produce offspring, which is represented in the graph as a self-recycling of $S$ and $F$, as well as transaction from $I$ to $S$. Some susceptible \textit{Daphnia} get infected by parasites, transitioning from $S$ to $I$. Both susceptible and infected \textit{Daphnia} contribute to the $R$ state upon death, releasing surviving parasites back into the environment. And the parasites continue to impact other individuals and get into another cycle. Food is consumed by both $S$ and $I$ individuals. Over time, elements in $F$ and $P$ also transit to the $R$ state.}
\label{fig:flow_SIRPF}
\end{figure}

@@

Each equation intricately ties the state variables to depict the complex dynamics of the single species Parasitized system. Building on the SIRPF model, the SIRPF2 model further adapts to a mixed-race Parasitized system by introducing another \textit{Daphnia} species. This advanced model incorporates both invasive($S_i , I_i$) and native($S_n , I_n$)hosts, along with food($F$) and parasites($P$), and delineates their specific interactions within the ecosystem as the following graph and model shows. In the SIRPF2 model, the subscript n is used to denote parameters and variables related to the native host, i for those pertaining to the invasive host, and p for the parasite spores. The SIRPF2 model thus offers a more detailed representation of the ecological dynamics by accounting for the complex interplay between different species and environmental factors in a mechanistic framework.
\begin{figure}
\begin{center}
%%%%% SEAIR diagram
\resizebox{8cm}{!}{
\begin{tikzpicture}[
  square/.style={rectangle, draw=black, minimum width=0.5cm, minimum height=0.5cm, rounded corners=.1cm, fill=blue!8},
  rhombus/.style={diamond, draw=black, minimum width=0.1cm, minimum height=0.1cm, fill=purple!8,aspect = 1},
  travel/.style={circle, draw=black, minimum width=0.5cm, minimum height=0.5cm, fill=green!8},
  report/.style={shape=regular polygon, regular polygon sides=8, draw, fill=red!8,minimum size=0.6cm,inner sep=0cm},
  bendy/.style={bend left=10},
  bendy2/.style={bend left=100},
  bendy3/.style={bend left=-100},
  >/.style={shorten >=0.25mm}, % redefine arrow to stop short of node
  >/.tip={Stealth[length=1.5mm,width=1.5mm]} % redefine arrow style
]
\tikzset{>={}}; % this is needed to implement the arrow redefinition

@@

\end{tikzpicture}
}
\end{center}
\vspace{-5mm}
\caption{A flow diagram for the SIRPF2 model. Apart from the quantified states above, the graph includes $R$ state, which represents deadth status. The flow diagram shows the how two species of \textit{Daphnia} interact with each other indirectly through $F$ and $P$. The marginal processes have been shown in SIRPF mode, and SIRPF2 model is the adaption form in mixed-species dynamcis}
\label{fig:flow_SIRPF2}
\end{figure}

@@

In the SIRPF2 model, detailed in Table 1, parameters and variables are distinctly annotated for native hosts (subscript $n$), invasive hosts (subscript $i$), and parasite spores (subscript $p$). Intraspecific competition among hosts is conceptualized via the Lotka-Volterra framework, with $r_n$ and $r_i$ denoting respective reproductive rates. Host density growth, influenced by algal density, is posited to be linear, indicating a direct relationship between food availability and \textit{Daphnia} reproduction. Mortality rates of susceptible hosts are denoted as $\theta_{S_n}$ and $\theta_{S_i}$. The model encapsulates the dynamics of infection through \textit{Metschnikowia bicuspidata} intake, affecting susceptible hosts. This interaction consequently modulates reproductive rates, represented as $\xi r_n$. Infection rates correlate with spore density ($P$) and host ingestion rates ($f_{S_n}$, $f_{S_i}$). Upon death, infected hosts ($\theta_{I_n}$, $\theta_{I_i}$) release spores, which are subsequently subject to ingestion and environmental degradation. The consumption dynamics of food by susceptible and infected hosts are captured by $\gamma_n$ and $\gamma_i$, and food growth is inversely proportional to its carrying capacity ($K_F$), reflective of resource limitation. The model also incorporates a constant removal rate $\delta$, representing the loss of hosts, spores, and food due to external sampling. The constraint of food supply, indicated by a constant rate $\mu = 0.37 \times 10^5$ cells per day, is based on observed declines in \textit{Daphnia} populations in the panel data, suggesting a pivotal role of food availability in driving population dynamics.\\

@@

\draw [dotted, thick, red] (3.5, -0.75) rectangle (6.5, 1.5);
\end{tikzpicture}
}
\end{center}
\vspace{-5mm}
\caption{A flow diagram for the SF and SF2 model. In this system, both types of \textit{Daphnia} and the food resource contribute to reproduction, depicted as self-recycling loops for $S_n$, $S_i$, and $F$. As time progresses, individuals in $F$, $S_n$, and $S_i$ transition to the $R$ state, signifying mortality. While the whole graph show SF2 model, if we remove the node in the dashed rectangle, the graph will represent the SF model.}
\label{fig:SF&SF2}
\end{figure}

@@

The SIRPF model is a powerful tool for analyzing ecological data, as it can be used to estimate parameters and predict future population dynamics. The model's flexibility allows for the incorporation of a variety of factors, including food resources and parasitism, that can be used to assess the impact of these factors on the population dynamics of \textit{Daphnia}. The model's ability to estimate parameters and predict future population dynamics is a powerful tool for analyzing ecological data. The model's flexibility allows for the incorporation of a variety of factors, including food resources and parasitism, that can be used to assess the impact of these factors on the population dynamics of \textit{Daphnia}.\\
